---
title: "Air Quality Monitoring Sites"
author: "BenS"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Air Quality Monitoring Sites}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

This vignette retrieves the current air-quality monitoring-site catalogue from
Environment Canterbury's public data website using `recanaq::get_stations()`.
The endpoint provides station identifiers, names, locations, airsheds, and the
latest observation time. It does not currently provide latitude and longitude,
so this workflow produces a searchable site catalogue rather than a map.

## Retrieve site data

```{r get-sites}
library(dplyr, warn.conflicts = FALSE)
library(knitr)
library(recanaq)

df_stations <- get_stations()
```

The returned data has one row per monitored site:

```{r inspect-sites}
str(df_stations)
```

## Prepare the catalogue

Add a date field and a simple freshness indicator. The threshold is deliberately
visible so it can be changed when reviewing current monitoring coverage.

```{r clean-sites}
recent_days <- 90

reference_date <- Sys.Date()

df_stations_clean <- df_stations |>
  mutate(
    latest_date = as.Date(latest_date_time),
    days_since_latest = as.integer(reference_date - latest_date),
    data_status = if_else(
      !is.na(days_since_latest) & days_since_latest <= recent_days,
      "Recent observation",
      "Older or missing observation"
    )
  ) |>
  arrange(city, site_name)
```

## Review the site catalogue

```{r site-table}
df_stations_clean |>
  select(
    site_no,
    site_name,
    long_name,
    city,
    air_shed,
    latest_date,
    data_status
  ) |>
  kable(
    format = "html",
    caption = paste("ECan monitoring sites retrieved on", reference_date)
  )
```

## Summarise coverage

The following summaries help identify how sites are distributed across cities
and airsheds, and which sites may need follow-up before analysis.

```{r airshed-summary}
df_stations_clean |>
  count(air_shed, sort = TRUE) |>
  kable(format = "html", caption = "Monitoring sites by airshed")
```

```{r freshness-summary}
df_stations_clean |>
  count(data_status, sort = TRUE) |>
  kable(format = "html", caption = "Latest-observation freshness")
```

## Save a copy

Save the retrieved catalogue when a reproducible snapshot is required. The
file records the response available from the website at the time of retrieval.

```{r export-sites}
output_dir <- "outputs"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

readr::write_csv(
  df_stations_clean,
  file.path(output_dir, paste0("ecan_monitoring_sites_", reference_date, ".csv"))
)
```
