---
title: "Theil-Sen Trend Analysis with ECan Data"
author: "Ben Scott"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Theil-Sen Trend Analysis with ECan Data}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette describes an end-to-end workflow for retrieving Environment
Canterbury (ECan) daily air-quality data and estimating PM10 and PM2.5 trends
with `openair::TheilSen()`.

The code is deliberately not evaluated during package builds. It requests live
remote data and can take several minutes to run, so execute the chunks
interactively when a current analysis is required. The original parameterized
Quarto report in `examples/theil_sen_trend_analysis.qmd` remains available for
scheduled or fully rendered analyses.

## Prerequisites

Install the optional analysis packages before running the workflow:

```{r packages}
install.packages(c("dplyr", "knitr", "lubridate", "openair", "readr", "tidyr"))
```

Load the package and analysis tools:

```{r libraries}
library(dplyr, warn.conflicts = FALSE)
library(lubridate, warn.conflicts = FALSE)
library(openair, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)
library(recanaq)
```

## Retrieve daily data

Choose the analysis period explicitly. Dates supplied to the package use
`dd/mm/yyyy` format.

```{r get-data}
from_date <- "01/01/2015"
to_date <- "31/12/2024"

df_final_long <- get_daily_all_stations(
  from_date = from_date,
  to_date = to_date
)
```

The result is a tidy, long-format data frame with one row per station, date,
and measurement.

## Prepare PM data

`openair` expects a `date` column and pollutant measurements in separate
numeric columns. Keep only the pollutants used in this analysis and reshape
the data into that format:

```{r data-cleaning}
pollutant_names <- c("PM10 (ug/m3)", "PM25 (ug/m3)")

df_daily_pm <- df_final_long |>
  filter(name %in% pollutant_names) |>
  pivot_wider(
    names_from = name,
    values_from = value
  ) |>
  rename(
    date = DateTime,
    site = StationName,
    PM10 = `PM10 (ug/m3)`,
    PM25 = `PM25 (ug/m3)`
  ) |>
  mutate(
    date = ymd(date, tz = "Pacific/Auckland"),
    month = month(date),
    year = year(date),
    days_per_month = days_in_month(date)
  )
```

Apply exclusions based on documented monitoring-site decisions. Review these
choices whenever sites move, close, or change purpose:

```{r site-decisions}
excluded_sites <- c("Washdyke Flat Road", "Waimate Stadium", "St Albans EP")

df_daily_pm_clean <- df_daily_pm |>
  filter(!site %in% excluded_sites)
```

## Check coverage

Review coverage by site and year before accepting trend results. Large gaps
may require a site-specific decision or a shorter analysis period.

```{r coverage}
coverage <- df_daily_pm_clean |>
  group_by(site, year) |>
  summarise(
    days_with_pm10 = sum(!is.na(PM10)),
    days_with_pm25 = sum(!is.na(PM25)),
    .groups = "drop"
  )

coverage
```

## Estimate Theil-Sen trends

The following calls estimate trends separately for PM10 and PM2.5, remove
seasonality, and require at least 75% data coverage:

```{r trend-pm10}
openair::TheilSen(
  df_daily_pm_clean,
  ylab = "PM10 (ug/m3)",
  pollutant = "PM10",
  deseason = TRUE,
  date.format = "%Y",
  type = "site",
  data.thresh = 75
)
```

```{r trend-pm25}
openair::TheilSen(
  df_daily_pm_clean,
  ylab = "PM2.5 (ug/m3)",
  pollutant = "PM25",
  deseason = TRUE,
  date.format = "%Y",
  type = "site",
  data.thresh = 75
)
```

Interpret the estimates alongside the coverage table and the site history.
Theil-Sen results are sensitive to the selected period, site exclusions, and
data completeness.

## Save analysis inputs

Save both the retrieved long-format data and the cleaned trend-analysis input
so the analysis can be audited without immediately repeating the remote
request:

```{r export}
output_dir <- "outputs"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
date_stamp <- Sys.Date()

readr::write_csv(
  df_final_long,
  file.path(output_dir, paste0("ecan_daily_long_", date_stamp, ".csv"))
)

readr::write_csv(
  df_daily_pm_clean,
  file.path(output_dir, paste0("ecan_pm_trend_input_", date_stamp, ".csv"))
)
```
